---
title: "Practical Machine Learning Course Project"
author: "Nadine Sanchez"
date: "August 6, 2017"
output: html_document
---

# Executive Summary

The goal of this project is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the excercise using the "classe" variable. 

The data used in this analysis is provided by http://groupware.les.inf.puc-rio.br/har .

## Loading the Data

This project requires both the pml-testing and pml-training csv data sets.

```{r}
if(!file.exists("pml-training.csv")){download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "pml-training.csv")}

if (!file.exists("pml-testing.csv")) {download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "pml-testing.csv")}

train_base <- read.csv('pml-training.csv', sep=',', stringsAsFactors = FALSE, strip.white = TRUE)
test_base <- read.csv('pml-testing.csv', sep=',', strip.white = TRUE, stringsAsFactors = FALSE)
```

## Set the seed

For reproducibilitiy purposes we will set the seed that we use in this project to be 300.

```{r}
set.seed(300)
```

## Load Libraries
```{r}
library(ggplot2)
library(caret)
library(randomForest)
library(rpart)
library(dplyr)
```

## Clean the datasset

We can remove unnecessary variables such as participant name, and variables that will not help answer the question.(This command utilizes the dplyr package)
```{r}
train <- train_base %>%
    select(-starts_with("avg_"),
           -starts_with("var_"),
           -starts_with("stddev_"),
           -starts_with("max_"),
           -starts_with("min_"),
           -starts_with("amplitude_"),
           -starts_with("kurtosis_"),
           -starts_with("skewness_"),
           -new_window,
           -num_window,
           -user_name,
           -raw_timestamp_part_1,
           -raw_timestamp_part_2,
           -cvtd_timestamp,
           -X)
dim(train)
```

The remaining dataset is composed of 53 measurements of the participants

## Splitting the Data

The dataset will be split such that a test set can be used to estimate the accuracy of the model. 70% of the data is used in the training set while 30% is used in the test set.

```{r}

#convert to factor variable
train$classe <- as.factor(train$classe)
```

```{r}
inTrain <- createDataPartition(y=train$classe, p=0.7, list=FALSE)
training <- train[inTrain, ]
testing <- train[-inTrain,]
```


## Building the Model

The following model uses a Random Forest algorithm and uses a 5-fold cross validation.

```{r}
fitControl           <- trainControl(method="cv", number=5, verboseIter=F)
modelForest          <- train(classe ~ ., data=training, method="rf", trControl=fitControl)


print(modelForest, digits=3)
```

```{r}
predictForest    <- predict(modelForest, newdata=testing)
confusionMatrix(testing$classe, predictForest)
```

### Conclusion

The accuracy of the random forrest algorithm is 99.13% . Therefore, the expected out-of-sample error is 0.87% . We can conclude that this is a good model.

## Making Predictions

The following commands are used to make predictions that are used for Part II of this assignment

```{r}
final_predictions      <- predict(modelForest, newdata=df_test_base)
print(final_predictions)
```